### 2021-12-07 lmcs/priority-gv.tex
top down polish

\subsection{Syntax of Types and Terms}
\subsubsection*{Session types}
Session types ($\ty{S}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{S}
  & \Coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]
Session types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ describe the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$. Types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ describe endpoints of a channel whose communication has finished, and over which we must synchronise before closing the channel. Each connective in a session type is annotated with a \emph{priority} $\cs{o}\in\mathbb{N}$.

\subsubsection*{Regular Types}
Types ($\ty{T}$, $\ty{U}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \Coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]
Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of \emph{pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$.

@@

Constants are best understood in conjunction with their typing and reduction rules in~\cref{fig:pgv-typing,fig:pgv-operational-semantics}.
Briefly, $\tm{\link}$ links two endpoints together, forwarding messages from one to the other, $\tm{\new}$ creates a new channel and returns a pair of its endpoints, and $\tm{\spawn}$ spawns off its argument as a new thread.
The $\tm{\send}$ and $\tm{\recv}$ functions send and receive values on a channel.
However, since the typing rules for PGV ensure the linear usage of endpoints, they also return a new copy of the endpoint to continue the session.
The $\tm{\close}$ and $\tm{\wait}$ functions close a channel.
We use syntactic sugar to make terms more readable: we write $\tm{\letbind{x}{M}{N}}$ in place of $\tm{(\lambda x.N)\;M}$, $\tm{\lambda\unit.M}$ in place of $\tm{\lambda z.\andthen{z}{M}}$, and $\tm{\lambda\pair{x}{y}.M}$ in place of $\tm{\lambda z.\letpair{x}{y}{z}{M}}$.
We recover $\tm{\fork}$ as $\tm{\lambda x.\letpair{y}{z}{\new\;\unit}{\andthen{\spawn\;{(\lambda\unit.x\;y)}}{z}}}$.

@@

While our configuration reduction is based on the standard evaluation for GV, the increased expressiveness of PGV allows us to simplify the relation on two counts.
\begin{enumerate*}[label=(\alph*)]
\item
\emph{We decompose the $\tm{\fork}$ construct}.
In GV, $\tm{\fork}$ creates a new channel, spawns a child thread, and, when the child thread finishes, it closes the channel to its parent. In PGV, these are three separate operations: $\tm{\new}$, $\tm{\spawn}$, and $\tm{\close}$. We no longer require that every child thread finishes by returning a terminated channel. Consequently, we also simplify the evaluation of the $\tm{\link}$ construct.
Intuitively, evaluating $\tm{\link}$ causes a substitution: if we have a channel bound as $\tm{\res{x}{y}{}}$, then $\tm{\link\;\pair{w}{x}}$ replaces all occurrences of $\tm{y}$ by $\tm{w}$. However, in GV, $\tm{\link}$ is required to return a terminated channel, which means that the semantics for $\tm{link}$ must \emph{create} a fresh channel of type $\ty{\tyends}/\ty{\tyendr}$. The endpoint of type $\ty{\tyends}$ is returned by the $\tm{link}$ construct, and a $\tm{\wait}$ on the other endpoint guards the \emph{actual} substitution. In PGV, evaluating $\tm{\link}$ simply causes a substitution.
\item
\emph{Our structural congruence is type preserving}. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well-typed configuration using $\equiv$.
\end{enumerate*}

@@

While our configuration typing is based on the standard typing for GV, it differs on two counts:
\begin{enumerate}[label= (\roman*) ]
\item \emph{we require that child threads return the unit value}, as opposed to a terminated channel; and
\item \emph{we simplify typing for parallel composition}.
\end{enumerate}
In order to guarantee deadlock freedom, in GV each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$. Consequently, associativity of parallel composition does not preserve typing. In PGV, we guarantee deadlock freedom using priorities, which removes the need for the channel pseudo-type $\ty{S^\sharp}$, and simplifies typing for parallel composition, while restoring type preservation for the structural congruence.

@@

  \begin{corollary}%
    \label{cor:pgv-closed-progress}
    If $\cseq[\phi]{\emptyenv}{\conf{C}}$, $\tm{\conf{C}}\centernot\cred$, and $\tm{\conf{C}}$ contains no endpoints, then $\tm{\conf{C}}\equiv\tm{\phi\;V}$ for some value $\tm{V}$.
  \end{corollary}
It follows immediately from \cref{thm:pgv-closed-progress-confs} and \cref{cor:pgv-closed-progress} that a term which does not return an endpoint will complete all its communication actions, thus satisfying deadlock freedom.

>>>>

\subsection{Syntax of Types and Terms}
\subsubsection*{Session types}
Session types ($\ty{S}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{S}
  & \Coloneqq & \ty{\tysend[\cs{o}]{T}{S}}
    \sep        \ty{\tyrecv[\cs{o}]{T}{S}}
    \sep        \ty{\tyends[\cs{o}]}
    \sep        \ty{\tyendr[\cs{o}]}
\end{array}
\]

Session types $\ty{\tysend[\cs{o}]{T}{S}}$ and $\ty{\tyrecv[\cs{o}]{T}{S}}$ describe the endpoints of a channel over which we send or receive a value of type $\ty{T}$, and then proceed as $\ty{S}$. Types $\ty{\tyends[\cs{o}]}$ and $\ty{\tyendr[\cs{o}]}$ describe endpoints of a channel whose communication has finished, and over which we must synchronise before closing the channel. Each connective in a session type is annotated with a \emph{priority} $\cs{o}\in\mathbb{N}$.

\subsubsection*{Types}
Types ($\ty{T}$, $\ty{U}$) are defined by the following grammar:
\[
\begin{array}{lcl}
  \ty{T}, \ty{U}
  & \Coloneqq & \ty{\typrod{T}{U}}
    \sep        \ty{\tyunit}
    \sep        \ty{\tysum{T}{U}}
    \sep        \ty{\tyvoid}
    \sep        \ty{\tylolli[\cs{p},\cs{q}]{T}{U}}
    \sep        \ty{S}
\end{array}
\]

Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of \emph{pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$.

@@

Constants are best understood in conjunction with their typing and reduction rules in~\cref{fig:pgv-typing,fig:pgv-operational-semantics}.

Briefly, $\tm{\link}$ links two endpoints together, forwarding messages from one to the other, $\tm{\new}$ creates a new channel and returns a pair of its endpoints, and $\tm{\spawn}$ spawns off its argument as a new thread.

The $\tm{\send}$ and $\tm{\recv}$ functions send and receive values on a channel.
However, since the typing rules for PGV ensure the linear usage of endpoints, they also return a new copy of the endpoint to continue the session.

The $\tm{\close}$ and $\tm{\wait}$ functions close a channel.

We use syntactic sugar to make terms more readable: we write $\tm{\letbind{x}{M}{N}}$ in place of $\tm{(\lambda x.N)\;M}$, $\tm{\lambda\unit.M}$ in place of $\tm{\lambda z.\andthen{z}{M}}$, and $\tm{\lambda\pair{x}{y}.M}$ in place of $\tm{\lambda z.\letpair{x}{y}{z}{M}}$.
We recover $\tm{\fork}$ as $\tm{\lambda x.\letpair{y}{z}{\new\;\unit}{\andthen{\spawn\;{(\lambda\unit.x\;y)}}{z}}}$.

@@

While our configuration reduction is based on the standard evaluation for GV, the increased expressiveness of PGV allows us to simplify the relation on two counts.
\begin{enumerate}[label=(\alph*)]
\item
\emph{We decompose the $\tm{\fork}$ construct}.
In GV, $\tm{\fork}$ creates a new channel, spawns a child thread, and, when the child thread finishes, it closes the channel to its parent. In PGV, these are three separate operations: $\tm{\new}$, $\tm{\spawn}$, and $\tm{\close}$. We no longer require that every child thread finishes by returning a terminated channel. Consequently, we also simplify the evaluation of the $\tm{\link}$ construct.

Intuitively, evaluating $\tm{\link}$ causes a substitution: if we have a channel bound as $\tm{\res{x}{y}{}}$, then $\tm{\link\;\pair{w}{x}}$ replaces all occurrences of $\tm{y}$ by $\tm{w}$. However, in GV, $\tm{\link}$ is required to return a terminated channel, which means that the semantics for $\tm{link}$ must \emph{create} a fresh channel of type $\ty{\tyends}/\ty{\tyendr}$. The endpoint of type $\ty{\tyends}$ is returned by the $\tm{link}$ construct, and a $\tm{\wait}$ on the other endpoint guards the \emph{actual} substitution. In PGV, evaluating $\tm{\link}$ simply causes a substitution.
\item
\emph{Our structural congruence is type preserving}. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well-typed configuration using $\equiv$.
\end{enumerate}

@@

While our configuration typing is based on the standard typing for GV, it differs on two counts:
\begin{enumerate}[label= (\roman*) ]
\item \emph{we require that child threads return the unit value}, as opposed to a terminated channel; and
\item \emph{we simplify typing for parallel composition}.
\end{enumerate}

In order to guarantee deadlock freedom, in GV each parallel composition must split \emph{exactly one} channel of the channel pseudo-type $\ty{S^\sharp}$ into two endpoints of type $\ty{S}$ and $\ty{\co{S}}$. Consequently, associativity of parallel composition does not preserve typing. In PGV, we guarantee deadlock freedom using priorities, which removes the need for the channel pseudo-type $\ty{S^\sharp}$, and simplifies typing for parallel composition, while restoring type preservation for the structural congruence.

@@

\begin{corollary}%
  \label{cor:pgv-closed-progress}
  If $\cseq[\phi]{\emptyenv}{\conf{C}}$, $\tm{\conf{C}}\centernot\cred$, and $\tm{\conf{C}}$ contains no endpoints, then $\tm{\conf{C}}\equiv\tm{\phi\;V}$ for some value $\tm{V}$.
\end{corollary}

An immediate consequence of \cref{thm:pgv-closed-progress-confs} and \cref{cor:pgv-closed-progress} is that \emph{a term which does not return an endpoint will complete all its communication actions, thus satisfying deadlock freedom.}
